A Pitman-Yor process prior for estimating shared evolutionary events

### 2021-02-25 manuscript/ms.tex
More work on the model description.

The primary goal is to infer the temporal clustering of divergences across
\ncomparisons comparisons.
At one extreme, all \ncomparisons could have diverged at the same time,
and at the other extreme, all \ncomparisons diverged independently.
This presents a model-choice problem, where the number of possible models
is the number of ways we can assign \ncomparisons comparisons to
$1, 2,  \ldots, \ncomparisons$ divergence times
\citep[the Bell number;][]{Bell1934}.
Below, we use \etimesets to represent the partitioning of comparisons to
divergence events, which we will also refer to as a ``divergence model,'' and
\etimes to represent the unique divergence times within \etimesets.

@@

\begin{linenomath}
\citet{Oaks2018ecoevolity} treated the number of divergence events and the
assignment of comparisons to those events as random variables under a Dirichlet
process \citep{Ferguson1973, Antoniak1974}.
The concentration parameter, \concentration, controls how clustered the
Dirichlet process is, and determines the probability of all possible \etimesets
(i.e., all possible set partitions of \ncomparisons comparisons).
To get these probabilities, we can assign comparisons to divergence events one
at a time, following a simple rule.
When assigning the $i^{th}$ comparison, we assign it to its own event
(i.e., a new divergence event with a unique time) with probability
\begin{equation}
    \frac{\concentration}{\concentration + i - 1},
    \label{eq:dpnewcat}
\end{equation}
or we assign it to an existing event $x$ with probability
\begin{equation}
    \frac{n_x}{\concentration + i - 1}.
    \label{eq:dpexistingcat}
\end{equation}
where $n_x$ is the number of comparisons already assigned to event $x$.
\end{linenomath}

@@

\subsection{Weighted-uniform (\wunif) model}
We have also implemented a uniform prior over divergence models, where we can
assume \emph{a priori} that every possible divergence model (\etimesets) is
equally probable.
Furthermore, we added a ``split weight'' parameter, which we denote as
\splitweight, to provide flexibility for the prior probabilities of
models with different numbers of events.
For a given model with \nevents divergence events, the relative probability
of each model with $\nevents + 1$ events is \splitweight,
and the relative probability of each model with $\nevents - 1$ events is
$\frac{1}{\splitweight}$.
More generally, the relative probability of each model with
$\nevents + n$
events is
$\splitweight{}^n$,
and the relative probability of each model with
$\nevents - n$
events is
$\frac{1}{\splitweight{}^n}$.
When $\splitweight = 1$, all divergence models are equally probable.

>>>>

The primary goal is to infer the temporal clustering of divergences across
\ncomparisons comparisons.
At one extreme, all \ncomparisons could have diverged at the same time,
and at the other extreme, all \ncomparisons diverged independently.
This presents a model-choice problem, where the number of possible models
is the number of ways we can assign \ncomparisons comparisons to
$\nevents = 1, 2,  \ldots, \ncomparisons$ divergence times
\citep[the Bell number;][]{Bell1934}.
Below, we use \etimesets to represent the partitioning of comparisons to
divergence events, which we will also refer to as a ``divergence model,'' and
$\etimes = \etime[1], \etime[2], \ldots, \etime[\nevents]$
to represent the times of the \nevents divergence events.

@@

\begin{linenomath}
\citet{Oaks2018ecoevolity} treated the number of divergence events and the
assignment of comparisons to those events as random variables under a Dirichlet
process \citep{Ferguson1973, Antoniak1974}.
The concentration parameter, \concentration, controls how clustered the
Dirichlet process is, and determines the probability of all possible \etimesets
(i.e., all possible set partitions of \ncomparisons comparisons).
To get the probability of any divergence model (\etimesets), we can assign
comparisons to divergence events one at a time, in any order, following a
simple rule.
When assigning the $i^{th}$ comparison, we assign it to its own event
(i.e., a new divergence event with a unique time) with probability
\begin{equation}
    \frac{\concentration}{\concentration + i - 1},
    \label{eq:dpnewcat}
\end{equation}
or we assign it to an existing event $x$ with probability
\begin{equation}
    \frac{n_x}{\concentration + i - 1}.
    \label{eq:dpexistingcat}
\end{equation}
where $n_x$ is the number of comparisons already assigned to event $x$.
The overall probability of the divergence model is then the product of each
comparison's probability of being added to its divergence event.
\end{linenomath}

@@

\subsection{Weighted-uniform (\wunif) model}
We also implemented what we are calling a ``weighted-uniform'' (\wunif) model
that has a single ``split weight'' parameter that controls the relative
probability of models with different numbers of events.
For a given model with \nevents divergence events, the relative probability
of each model with $\nevents + 1$ events is \splitweight,
and the relative probability of each model with $\nevents - 1$ events is
$\frac{1}{\splitweight}$.
More generally, the relative probability of each model with
$\nevents + n$
events is
$\splitweight{}^n$,
and the relative probability of each model with
$\nevents - n$
events is
$\frac{1}{\splitweight{}^n}$.
Models with the same number of divergence events
are always equally probable,
and
when $\splitweight = 1$, all possible divergence models (\etimesets) across all
numbers of events are equally probable.

\begin{linenomath}
To formalize this as a proper probability mass function,
the probability of any divergence model is
\begin{equation}
    p(\etimesets \given \splitweight, \ncomparisons, \nevents) = 
    \frac{
        \splitweight^{(\nevents - 1)}
    }{
        \sum\limits_{k=1}^{k=\ncomparisons} S_2(\ncomparisons, k) \splitweight^{(k-1)}
    },
    \label{eq:uniformdivmodelprob}
\end{equation}
where $S_2(\ncomparisons, k)$ is the Stirling number of the second kind which
is equal to the number of ways of assigning the \ncomparisons comparisons to
$k$ divergence events.
\end{linenomath}
